---
title: "TCGA_BRCA"
author: "Neal Liu"
date: "2024-01-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}

library(TCGAbiolinks)
library(RUVIIIPRPS)
library(MOFA2)
library(ggplot2)
library(dplyr)
library(edgeR)
library(parallel)
library(readr)
library(ggExtra)
library(cowplot)
library(SummarizedExperiment)
library(biomaRt)
library(umap)
library(Rtsne)
library(singscore)

source('Helper_functions.R')

library(devtools)
install_github('RMolania/RUVIIIPRPS' , force = T)

```


```{r}

data_transcriptome <- readRDS('Data/BRCA (Breast invasive carcinoma).rds')
data_methylation <- readRDS('Data/BRCA methylation illumina 27k.rds')
data_methylation_450k <- readRDS('Data/BRCA Methylation 450k.rds')


```

```{r}

sum(edgeR::filterByExpr(assay(data_transcriptome,'HTseq_counts'), min_count = 10, min.total.count = 80))

p1 <- plot_PCA(log2(assay(data_transcriptome, 'HTseq_FPKM')+1)[edgeR::filterByExpr(assay(data_transcriptome,'HTseq_counts'), min_count = 10, min.total.count = 80),],'fpkum uq transcriptome', data_transcriptome@colData$BatchId_mda, 'batch')

p2 <- plot_PCA(log2(assay(data_transcriptome, 'HTseq_FPKM')+1)[edgeR::filterByExpr(assay(data_transcriptome,'HTseq_counts'), min_count = 10, min.total.count = 80),],'fpkum uq transcriptome', data_transcriptome@colData$subtype_BRCA_Subtype_PAM50__RNAseq, 'subtype')

filtered_transcriptome <- data_transcriptome[edgeR::filterByExpr(assay(data_transcriptome,'HTseq_counts'), min_count = 10, min.total.count = 80),]

p1


```
```{r}
p2

```

```{r}

plot_3D_PCA(log2(assay(data_transcriptome, 'HTseq_FPKM')+1)[edgeR::filterByExpr(assay(data_transcriptome,'HTseq_counts'), min_count = 10, min.total.count = 80),], data_transcriptome@colData$subtype_BRCA_Subtype_PAM50__RNAseq)

```


```{r, fig.width=19, fig.height= 6}
plot_RLE(log2(assay(data_transcriptome,'HTseq_counts')+1)[edgeR::filterByExpr(assay(data_transcriptome,'HTseq_counts'), min_count = 10, min.total.count = 80),1:500], data_transcriptome@colData$BatchId_mda[1:500], ylimit = c(-4, 4))

```

```{r}
data_methylation <- data_methylation[which(apply(assay(data_methylation), 1, function(x) {sum(is.na(x))/length(x)}) <= 0.2), ]
assay(data_methylation) <- replace_all_NAs_with_row_means(assay(data_methylation))
assay(data_methylation, 'M_values') <- log2(assay(data_methylation)/(1-assay(data_methylation)))

plot_PCA(assay(data_methylation,'M_values'),'M values', data_methylation@colData$paper_BRCA_Subtype_PAM50, 'pam50 subtype')

```

```{r, fig.width=19, fig.height= 6}
plot_RLE(assay(data_methylation,'M_values'),data_methylation@colData$site_of_resection_or_biopsy, ylimit = c(-4,4))


```

```{r}

# remove methylations that have more than 10% NAs and methylations that do not map to a gene. this ends up with around 330k rows.
data_methylation_450k <- data_methylation_450k[which(apply(assay(data_methylation_450k), 1, function(x) {sum(is.na(x))/length(x)}) <= 0.1), ]
data_methylation_450k <- data_methylation_450k[!is.na(data_methylation_450k@rowRanges@elementMetadata@listData$gene),]

# replace remaining NAs with row means and calculate M values
assay(data_methylation_450k) <- replace_all_NAs_with_row_means(assay(data_methylation_450k))

assay(data_methylation_450k, 'M_values') <- log2(assay(data_methylation_450k)/(1-assay(data_methylation_450k)))

methylation_450k_pca <- list(BiocSingular::runSVD(t(assay(data_methylation_450k,'M_values')),k=10,center=T))[[1]]

plot_PCA(methylation_450k_pca,'M values', data_methylation_450k@colData$paper_BRCA_Subtype_PAM50, 'pam50 subtype', is_pca_obj = T, pcs = c(1,2))

```

Since the methylations are split into 27k and 450k, both having a significant amount of patients, 
it might be a good idea to combine them by keeping common sites between both. However, we must first 
check if there is significant effects between the two platforms because we don't know if they use 
different reagents etc which will have effects on the data distribution. 


```{r}

common_methylations <- intersect(rownames(assay(data_methylation,'M_values')),rownames(assay(data_methylation_450k,'M_values')))
common_patients <- intersect(data_methylation_450k@colData$patient, data_methylation@colData$patient)

`27k_test_matrix` <- assay(data_methylation,'M_values')[common_methylations, sort(which(data_methylation@colData$patient %in% common_patients))]
`450k_test_matrix` <- assay(data_methylation_450k,'M_values')[common_methylations, sort(which(data_methylation_450k@colData$patient %in% common_patients))]

methylation_platform_correlations <- as.matrix(unlist(lapply(1:length(common_methylations), 
                                                             function(x) {stats::cor(`27k_test_matrix`[x,],
                                                                              `450k_test_matrix`[x,], 
                                                                              method='pearson')})))

p1 <- ggplot(mapping = aes(x= `27k_test_matrix`[,1], y= `450k_test_matrix`[,1])) +
  geom_point( alpha = 0.05) +
  theme_minimal() +
  labs(x = '27k (vial A)', y = '450k (vial B)')+
  ggtitle('M values of patient A0DC')+
  theme(panel.border=element_rect(colour = "grey80", fill=NA, size=0.8),
          aspect.ratio = 1/1.05,
          axis.line = element_line(colour = "grey75", linewidth = 1.1),
          panel.grid.major = element_line(color = "grey96"),)

# 0.9489513 correlation

p2 <- ggplot(mapping = aes(x= assay(data_methylation,1)[common_methylations, "TCGA-A7-A0DC-01A-11D-A00Y-05"], 
                     y= assay(data_methylation_450k,1)[common_methylations, "TCGA-A7-A0DC-01B-04D-A22R-05"])) +
  geom_point( alpha = 0.05) +
  theme_minimal() +
  labs(x = '27k (vial A)', y = '450k (vial B)')+
  ggtitle('Beta values')+
  theme(panel.border=element_rect(colour = "grey80", fill=NA, size=0.8),
          aspect.ratio = 1/1.05,
          axis.line = element_line(colour = "grey75", linewidth = 1.1),
          panel.grid.major = element_line(color = "grey96"),)

# 0.9682381 correlation 

gridExtra::grid.arrange(p1,p2, ncol=2)

```

```{r}

p1 <- ggplot(mapping = aes(x= assay(data_methylation,2)[common_methylations, "TCGA-BH-A18V-01A-11D-A12E-05"], 
                     y= assay(data_methylation_450k,2)[common_methylations, "TCGA-BH-A18V-06A-11D-A212-05"])) +
  geom_point( alpha = 0.05) +
  theme_minimal() +
  labs(x = '27k (primary tumour)', y = '450k (metastatic)')+
  ggtitle('M values of patient A18V')+
  theme(panel.border=element_rect(colour = "grey80", fill=NA, size=0.8),
          aspect.ratio = 1/1.05,
          axis.line = element_line(colour = "grey75", linewidth = 1.1),
          panel.grid.major = element_line(color = "grey96"),)

# 0.9138258 correlation

p2 <- ggplot(mapping = aes(x= assay(data_methylation,1)[common_methylations, "TCGA-BH-A18V-01A-11D-A12E-05"], 
                     y= assay(data_methylation_450k,1)[common_methylations, "TCGA-BH-A18V-06A-11D-A212-05"])) +
  geom_point( alpha = 0.05) +
  theme_minimal() +
  labs(x = '27k (primary tumour)', y = '450k (metastatic)')+
  ggtitle('Beta values')+
  theme(panel.border=element_rect(colour = "grey80", fill=NA, size=0.8),
          aspect.ratio = 1/1.05,
          axis.line = element_line(colour = "grey75", linewidth = 1.1),
          panel.grid.major = element_line(color = "grey96"),)

# 0.9146544 correlation 

gridExtra::grid.arrange(p1,p2, ncol=2)

```

```{r}

data_methylation_450k@colData <- data_methylation_450k@colData[colnames(data_methylation_450k@colData) %in% colnames(data_methylation@colData)]
data_methylation_450k@colData$platform <- '450k'


data_methylation@colData <- data_methylation@colData[match(colnames(data_methylation_450k@colData),colnames(data_methylation@colData))]
data_methylation@colData$platform <- '27k'

illumina_27k_rowdata <-  data_methylation@rowRanges
illumina_450k_rowdata <-  data_methylation_450k@rowRanges
rowRanges(data_methylation) <- NULL
rowRanges(data_methylation_450k) <- NULL

```
  

```{r}
data_methylation_complete <- SummarizedExperiment(assays = list( beta_values  = cbind(assay(data_methylation_450k[common_methylations,],1), 
                                                                                      assay(data_methylation[common_methylations,],1)),
                                                                 M_values = cbind(assay(data_methylation_450k[common_methylations,],2), 
                                                                                      assay(data_methylation[common_methylations,],2))),
                                                  rowData = illumina_450k_rowdata[common_methylations,],
                                                  colData =  rbind(data_methylation_450k@colData, data_methylation@colData),
                                                  )

```


```{r}

all_methylation_pca <- run_PCA(assay(data_methylation_complete,'M_values'))

plot_PCA(all_methylation_pca,'M values of all methylation patients', data_methylation_complete@colData$platform, 'platform', is_pca_obj = T)


```

```{r}
plot_PCA(all_methylation_pca,'M values of all methylation patients', data_methylation_complete@colData$paper_BRCA_Subtype_PAM50 ,'subtypes', is_pca_obj = T)

```

```{r}
plot_3D_PCA(all_methylation_pca$u, data_methylation_complete@colData$paper_BRCA_Subtype_PAM50, is_PCA_obj = T)

```

```{r}
filtered_transcriptome@colData <- t(remove_NA_rows(t(data_transcriptome@colData)))

filtered_transcriptome_matched_samples <- filtered_transcriptome[, which(filtered_transcriptome@colData$sample__RNAseq %in% intersect(data_methylation_complete@colData$sample, filtered_transcriptome@colData$sample__RNAseq))]
data_methylation_complete_matched_samples <- data_methylation_complete[, which(data_methylation_complete@colData$sample %in% intersect(data_methylation_complete@colData$sample, filtered_transcriptome@colData$sample__RNAseq))]

```

```{r, fig.width=19, fig.height= 7}
plot_RLE(assay(data_methylation_complete_matched_samples,'M_values')[, 700:1000], data_methylation_complete_matched_samples@colData$platform[700:1000], ylimit = c(-5,5))

```
```{r}
data_methylation_complete_matched_samples@colData$Plate <- substr(data_methylation_complete_matched_samples@colData$barcode, 22,25)


```

```{r, fig.width=19, fig.height= 7}
plot_RLE((assay(data_methylation_complete_matched_samples,'M_values')[,which(data_methylation_complete_matched_samples@colData$platform == '450k')]), (data_methylation_complete_matched_samples@colData$Plate[which(data_methylation_complete_matched_samples@colData$platform == '450k')]), ylimit = c(-4,4))

```


```{r}
table(data_methylation_complete_matched_samples@colData$Plate)
table(filtered_transcriptome_matched_samples@colData$BatchId_mda)

data_methylation_complete_matched_samples <- data_methylation_complete_matched_samples[,!duplicated(data_methylation_complete_matched_samples@colData$sample)]
data_methylation_complete_matched_samples<- data_methylation_complete_matched_samples[,match(filtered_transcriptome_matched_samples@colData$sample__RNAseq, data_methylation_complete_matched_samples@colData$sample)]
data_methylation_complete_matched_samples@colData$singscore_purity <- filtered_transcriptome_matched_samples@colData$Purity_singscore

```


```{r, fig.width= 9}

methylation_normalized <- data_methylation_complete_matched_samples
methylation_normalized@colData$Plate <- factor(methylation_normalized@colData$Plate)
methylation_normalized@colData$platform <- factor(methylation_normalized@colData$platform)
methylation_normalized@colData$subtype <- filtered_transcriptome_matched_samples@colData$subtype_BRCA_Subtype_PAM50__RNAseq
assay(methylation_normalized,2) <- as.matrix(assay(methylation_normalized,2))

methylation_normalized <- RUVIIIPRPS::supervisedPRPS(methylation_normalized[,which(!is.na(methylation_normalized@colData$subtype))], 'M_values',
                                                     bio.variable = 'subtype',
                                                     uv.variables = c('platform','Plate','singscore_purity'), 
                                                     batch.variable = 'Plate', 
                                                     min.sample.for.prps = 3,
                                                     min.sample.per.batch = 3,
                                                     apply.log = F,
                                                     pseudo.count = 1,
                                                     assess.se.obj = F,
                                                     assess.cor.variables = T,
                                                     save.se.obj = TRUE,
                                                     verbose = T)


# plotting prps map

prps_map <- as.data.frame(methylation_normalized@colData)[,c('Plate','paper_BRCA_Subtype_PAM50')] %>%
    dplyr::count(Plate, paper_BRCA_Subtype_PAM50)

prps_map$use <- ifelse(prps_map$n > 2, 'yes','no')

ggplot(prps_map, aes(x = Plate, y = paper_BRCA_Subtype_PAM50)) +
  geom_count(aes(color = use)) +
  geom_text(aes(
    label = n,
    hjust = 0.5,
    vjust = 0.5
  )) +
  xlab('Batch') +
  ylab('Confident subtypes') +
  theme_bw() +
  theme(axis.line = element_line(colour = 'grey35', size = 1.1),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 0),
    axis.text.x = element_text(size = 10,angle = 45,hjust = 1),
    axis.text.y = element_text(size = 12, angle = 45, hjust = 1),
    legend.position = 'none',
    panel.grid.major = element_line(color = "grey95"),
    aspect.ratio = 1/2)

```

```{r}
methylation_normalized <- RUVIIIPRPS::supervisedFindNCG(methylation_normalized, 'M_values',
                                                     bio.variable = 'subtype',
                                                     uv.variables = c('platform','Plate','singscore_purity'), 
                                                     no.ncg = 1000,
                                                     apply.log = F,
                                                     assess.se.obj = F)


```

```{r}
NCGs <- methylation_normalized@metadata$NCG

pca_NCGs <- run_PCA(assay(methylation_normalized,2)[NCGs, ])

plot_3D_PCA(pca_NCGs$u, methylation_normalized@colData$platform, is_PCA_obj = T)


```

```{r}
plot_3D_PCA(pca_NCGs$u, methylation_normalized@colData$Plate, is_PCA_obj = T)

```

```{r}

ggplot(mapping = aes(x=methylation_normalized@colData$singscore_purity, y=(pca_NCGs$u[,1])))+
  geom_point(alpha=0.62, color= 'grey30') + # alpha for transparency to see overlap
  labs(y="PC1 of NCGs", 
       x="Purity") +
  theme_minimal()+
  ggtitle('1st PC of negative control genes vs purity') +
  theme(panel.border=element_rect(colour = "grey80", fill=NA, size=0.8),
        aspect.ratio = 1/1.05,
        axis.line = element_line(colour = "grey75", linewidth = 1.1),
        panel.grid.major = element_line(color = "grey96"),)


```

```{r}
replicate.data=t(do.call(cbind, methylation_normalized@metadata[['PRPS']][['supervised']]))

methylation_normalized <- RUVIIIPRPS::ruvIIIMultipleK(se.obj = methylation_normalized, 
                                                      assay.name = 'M_values',
                                                      apply.log = F,
                                                      k = 10,
                                                      ctl= methylation_normalized@metadata[['NCG']],
                                                      assess.se.obj = F, 
                                                      replicate.data = replicate.data,
                                                      save.se.obj = T,
                                                      remove.na = 'measurements')


```


```{r}
methylation_normalized <- RUVIIIPRPS::normalise(se.obj = methylation_normalized,
                                                         assay.name='M_values',
                                                         apply.log = F, 
                                                         k=12, 
                                                         bio.variable.prps = 'subtype', 
                                                         uv.variables.prps = c('singscore_purity','Plate','platform'),
                                                         batch.variable.prps = 'Plate',
                                                         min.sample.per.batch.prps = 3,
                                                         norm.assay.name.ncg = 'M_values',
                                                         bio.variables.ncg = 'subtype',
                                                         uv.variables.ncg = c('singscore_purity','Plate','platform'),
                                                         assess.se.obj = F
                                                         )
```









